```{r}
#| echo: false

source("_setup.R")
source("_objects.R")
```

# Publication design {#sec-publication-table}

Tables are common is scientific publications but usually rather minimalist design is preferred. Such tables are not only easy to read but also easy to create with the `{gt}` package - a perfect occasion for the first table of this tutorial. For the demonstration purposes, we will summarize δ^13^C measurement by families and countries.

```{r}
#| label: summarize-by-family-country

veg_summarized <- dat_vegetation |> 
    group_by(family, country, type) |>
    summarize(
      n_plants = n(), # number of underlying measurements
      mean_d = mean(delta),
      sd_d = sd(delta),
      ) |> 
    ungroup()
```

## Simplest table

Displaying a data frame as a `gt` table is a straightforward process and the default design is pleasing. However, our table will require some additional formatting to satisfy taxonomical and scietific conventions given by the data.

```{r}
#| label: minimal-gt-code

tbl_plain <- veg_summarized |> 
  gt() |> 
  tab_options(table.width = px(800))
```

```{r}
#| label: tbl-minimal-code
#| echo: false

tbl_plain
```


## Scientific and  other formatting

We will choose if there are any columns we would like to hide - here the `n_plants` which we introduced as helper column to control conditional formatting (in @sec-conditional-rows). We take advantage of `sub_missing()` function which helps us to handle missing values (`NA`s) according to our preference without the need to modify the input data directly or by introducing a helper column.

With `tab_style()` we can easily italicize an entire column and with `cols_label()` we can turn existing variable names into a meaningful and appealing column labels. When setting up a spanner over the numerical columns, we use `markdown` to achieve proper formatting. `fmt_number()` makes common number formatting options a simple task - we use to control number of decimal places.


```{r}
#| label: code-tbl-publication

tbl_formatted <- tbl_plain |>
    # this is only a helper column used for conditional formatting and it should not be displayed in the final table
    cols_hide("n_plants") |>
    # replace default NA values with a "-"
    # Note: modifying the data with {dplyr} functions would work equally well but {gt} has a devoted function for this purpose
    sub_missing(columns = "sd_d", missing_text = "-") |>
    tab_style(style = list(
        cell_text(style = "italic")),
    # family names in italics
    locations = cells_body(columns = "family")
    ) |> 
    # ^13^ is markdown for superscript, &delta is html for lower case Greek delta
    tab_spanner(columns = contains("_d"), label = md("&delta;^13^C")) |> 
    cols_label(
        family = "Family",
        country = "Country",
        type = "Type",
        mean_d = "Mean", 
        sd_d = "SD"
    ) |> 
    fmt_markdown(columns = c("type")) |> 
    fmt_number(columns = c("mean_d", "sd_d"), decimals = 2)
```

```{r}
#| label: tbl-publication
#| echo: false

tbl_formatted |>
  tab_options(table.width = px(800)) |> 
  cols_width(matches("country") ~ px(200))
```


## Custom theme

A publication may contain multiple figures. Instead of repeating the same code again and potentially making the same adjustments at multiple places if we decide to make changes to the table design, we can define a custom theme once. We use `{gt}` function and formatting options as with a regular table but wrap all of our code into a function. In our theme, we will define font family and size for individual components of the table. We will also ensure that column headings are displayed in bold font face.

```{r}
#| label: gt-publication-theme

paper_gt_theme <- function(gt_table) {
  gt_table |> 
    # overall aspects such as font and padding
    tab_options(
      table.font.names = "Times New Roman",
      data_row.padding = px(6),
      heading.align = "center",
      heading.title.font.size = px(26),
      heading.subtitle.font.size = px(14),
      table_body.hlines.width = px(0)
    ) |> 
    # column labels and spanners in bold font face
    tab_style(
        style = cell_text(weight = "bold"),
        locations = cells_column_labels(everything())
    ) |> 
    tab_style(
        style = cell_text(weight = "bold"),
        locations = cells_column_spanners(spanners = everything())
    )      
}
```


::: {.callout-warning}
## Custom theme placement

The custom scheme contains formatting options for column spanners and the theme definition **must** thus come only once spanner were defined. In other case, they will remain unaffected by the custom theme.
:::

```{r}
#| label: tbl-publication-themed

tbl_formatted |>
  # Note: we intorduced column spanners earlier. It is important that the custom theme is used after a (all) spanner(s) were created because these are targeted by the theme
  paper_gt_theme()|>
  tab_options(table.width = px(800)) |> 
  cols_width(matches("country") ~ px(200))
```


## Conditional row formatting {#sec-conditional-rows}

We take the existing, formatted table a little further. In some families, the number of measured species was low and we will use this information for conditional formatting of the affected rows. We will also include a table footnote to explain the formatting right in place.

We stored the previous table as a gt table and we will take advantage of it. We can add additional formatting options to existing gt object which is more elegant and less error-prone than copy-pasting of longer pieces of code.

We take advantage of `n_plants` "helper column" to grey out families with less than three measurements, with `tab_footnote()` we can easily provide more details about meaning of the formatting options.

```{r}
#| label: code-tbl-publication-ready-conditional

tbl_publication_conditional <- tbl_formatted |>
  paper_gt_theme() |>
  # families with less than three measurements in grey
  tab_style(style = list(
    cell_text(color = "grey50")
    ),
    locations = cells_body(columns = everything(),
                           rows = n_plants < 3)
    ) |> 
  # conditional formatting was introduced and it should be explained - a table footnote is a good place to do so
  tab_footnote(footnote = md(paste(folio_footnote, "<br>", 
                                   "Families which have less than three measurements are greyed out."))) |> 
  tab_style(style = cell_text(align = "left"), locations = cells_footnotes())
```


```{r}
#| label: tbl-publication-ready-conditional
#| echo: false

tbl_publication_conditional |>
  tab_options(table.width = px(800)) |> 
  cols_width(matches("country") ~ px(200))
```


## Default table processing by Quarto

::: {.callout-tip}
## Table processing in Quarto

Quarto, by default, applies a Bootstrap CSS class (`table-striped`) to tables generated from code chunks, which introduces row stripping in HTML tables. Row stripping is not common in publications and we thus turn it off in the code chunk below (with `html-table-processing: "none"`) to show how a final publication-ready table might look like.
:::

```{r}
#| label: tbl-publication-ready-conditional-stripped
#| html-table-processing: "none"
#| echo: false

tbl_publication_conditional |>
  tab_options(table.width = px(800)) |> 
  cols_width(matches("country") ~ px(275))
```




